---
title: "Getting started"
author: "Abas Shkembi"
date: "August 24, 2026"
format: html
---

::: callout-note
For those without a background in using R, this page will contain a semi-exhaustive list of important tidbits about the R language that will be useful to know for this class. Some of these topics will be visited and revisted throughout class lectures, but we wanted a central location for some of the key features of R you should know.
:::

## Before we get started

-   R is a very flexible language there is often more than one way to complete a task
-   The `tidyverse` packages have their own spin on the R language
-   What we present is the best way we think to write R code to complete tasks
-   You might find a way that works better for you

### General Programming Tips

-   Write everything in your code file
-   Fix mistakes, but don’t delete code
    -   You can always use the ‘#’ to comment out the code so it does not run
-   You can undo mistakes with Ctrl-Z
-   Recycle your code (by copying and pasting)
    -   You can often use the same code and just tweak the column names
-   Take your time
-   Speak (or write) out what you want your code to do in English
-   Refer to the example code
-   You are going to make mistakes, that’s expected

## What is R?

R is a computer language. R is primarily used for statistical analysis and graphics development, including but not limited to:

-   data handling and manipulation
-   data analysis
-   data visualization

There are many other statistical programs, such as STATA, SAS, and SPSS. What differentiates R from other statistical programs is that R is a *language*, not a *program*. Wherein STATA you program your computer to complete a set of pre-determined commands, you can develop your own set of commands in R from scratch.

This is because much of R is essentially an interpreter of C, C++, and Fortran, making its uses much more flexible and varied than other statistical programs. In programs like STATA, you are limited to the set of commands already made available to you. Meanwhile, if a particular command does not exist in R, you can create it yourself.

This is because R is a *functional* language. In other words, nearly all of the commands in R can be performed using some function `f(x)`:

```{r}
sum(c(0, 1, 2, 3))
```

where in this instance, `f()` is `sum()`, and `x` is the vector `c(0, 1, 2, 3)`. However, if the `sum()` function did not exist, you could create your own:

```{r}
# initialize that you want to create a function called `my_sum()`
my_sum <- function(x) {
  # Initialize the sum to 0
  total <- 0
  
  # Iterate through the vector and add each element to the total
  for (val in x) {
    total <- total + val
  }

  # Return the final sum
  return(total)
}

my_sum(c(0, 1, 2, 3))
```

This is part of what separates R from other statistical programs. Anyone can create functions, and as a result, R is a composite of *packages* (i.e., set of related functions) that many independent developers created and made freely available to anyone with a computer and internet. This makes R *open source*. Meanwhile, programs like STATA and SAS are examples of *closed source* programs, meaning that the code is owned and modified by STATA and SAS developers only. This is why R is free to use, while STATA and SAS require paid licenses.

> Note! There are advantages and disadvantages to open-sourced programs. R is much more diverse as a result of its thousands of packages; however, it is up each developer to ensure their code and packages work correctly. Programs like STATA is less diverse due to being closed-source; however, STATA ensures that its code works as intended.

R is also a *vectorized* language. This means that operations are applied to entire vectors at once, rather than element by element. What this means in practice is that R is much more efficient when it is analyzing vectors. For example, say you want to know element-by-element the sums of two vectors: `c(1, 2, 3)` and `c(3, 2, 1)`. Whereas many other languages require looping through the `k=3` elements of each vector:

```{r}
vector_sum <- c()
for(i in 1:3){
  vector_sum[i] <- c(1, 2, 3)[i] + c(3, 2, 1)[i]
}
print(vector_sum)
```

you can simply add two vectors in R:

```{r}
c(1, 2, 3) + c(3, 2, 1)
```

Understanding that R is a vectorized language is not important if you work with small-to-medium data. However, when analyzing big data, this point is critical. This is because every form of data in R (e.g., a vector, list, data frame, tibble) is stored on your own computer as an array of vectors and is read from top-to-bottom then left-to-right. Thus, each column in your dataset is a vector, making it much more efficient to operate on a column than on a row.

## What is RStudio?

While R is a computer language, RStudio is an *Integrated Development Environment* (IDE) that provides you with a more user-friendly interface when coding in R. RStudio allows you to write a `.r` script that can execute that R code, debug it, find typos, provide you with a window for plots you generate, and display all of the data types stored in your current environment.

> Note! It is not necessary to use RStudio to run R code; however, RStudio will make it much easier to develop your R code.

## Prerequisites to Code

Before you start diving into manipulating and analyzing code, there are some basic prerequisites you need to master:

-   Opening an R script
-   Setting your working directory
-   Loading a data file
-   Installing and loading packages

### Opening an R script

While you can code using the R console, using an R script provides you with the ability to save your code, quickly edit and debug any errors, and reproduce similar analyses in the future. You'll notice you have a few options when opening a new file in R, such as an R script, Quarto document, RMarkdown file, and many more. For simplicity, begin using an R script, which will open the simplest blank R file that will allow you to execute commands in a sequential manner.

### Setting your working directory

All files in your computer can be accessed using different file paths. Your working directory is the file path that you point your computer towards when reading documents. It is good coding practice to create a new folder for a particular coding project that will house all documents related to your analysis (e.g., datasets, code, figures created, etc.). You want to specify this particular project folder as your working directory to make it easier when loading in or creating new files while coding.

When you first open R, your working directory is your entire computer, as you have not yet specified a particular folder. You can specify this folder using the set working directory command (`setwd()`). For example, on a Apple Macbook, let's say my username is `industrialhygiene`, and I have a folder under the file path `Documents > Coding > EHS655` such that "EHS655" is the main folder name where I will be doing my R analysis. Then, I can set the working directory using the following code:

```{r, eval=F}
setwd("/Users/industrialhygiene/Documents/Coding/EHS655")
```

### Loading a file

Now, if I read in a dataset file (e.g., "noise_data.csv"), I can simply call the dataset filename directly since my RStudio already knows we are in the "EHS655" folder:

```{r, eval=F}
read.csv("noise_data.csv")
```

Otherwise, if I did not set my working directory already, I would've needed to read the file using the entire file path, since my RStudio would not have understood what folder to look in:

```{r, eval=F}
read.csv("/Users/industrialhygiene/Documents/Coding/EHS655/noise_data.csv")
```

This approach works, but it is simpler and more ideal for code reproducibility to store all your project files in a single folder and set your working directory to that folder.

### Installing and loading packages

All functions that are stored in R to run a particular command are stored in packages. Already initialized in your RStudio when you first open it are "base R" packages, such as the `base`, `datasets`, and `utils` packages. If you want to install a new package to use additional commands, you will have to install and load them into your environment manually. This is unlike other licensed statistical programs, such as STATA, which are already pre-loaded with all of the possible commands as soon as you open the program.

For example, base R does not have the ability to read excel files. In order to do this, you will have to first install the `xlsx` package.

```{r, eval=F}
install.packages("xlsx")
```

However, you have not yet loaded the package into your R environment so that you can use its commands. At this point, you have simply downloaded the files into your own computer so that the pacakge can be used in the future. To load the package, you use the `library()` function.

```{r, eval=F}
library("xlsx")
```

Now that you have run this code, you are ready to use the functions in the `xlsx` package. You can quickly view the available functions in RStudio by typing `xlsx::`. The `::` tell your R console to look at the particular package only. Otherwise, you can search package documentation on the internet to look up the particular functions in your package

> Note! You will have to run the "library()" command every time you open your RStudio to use the loaded functions.

Finding new R packages to download will come naturally from searching online for new coding techniques. Over time, there will be some packages you notice that you use over and over again, and you will get into the habit of always loading in these packages as soon as you open a new R script.

## Base R Code

The code presented below is just a primer of the main syntax, operations, and data types of R. It is not exhaustive. There are many tutorials and books online that provide in-depth details on this topics.

### Basic syntax

You can output text using single or double quotes:

```{r}
'Hello SPH!'
```

You can output numbers by just typing the number:

```{r}
8
```

### Basic math

You can perform basic mathematical operations:

```{r, results='hide'}
2+2
2-2
2*2
2/2
2^2
```

### Variables

You can assign and store a variable or data type using the assignment operator `<-`, `->`:

```{r, results='hide'}
# a vector named x
x <- 1
```

You can print this variable by just directly calling it or using the `print()` function:

```{r}
x
```

### Booleans

A form of logical values (i.e., `TRUE` or `FALSE`) using:

-   `>`: greater than
-   `>=`: greater than or equal to
-   `<`: less than
-   `<`: less than or equal to
-   `==`: equal to
-   `!=`: not equal to

```{r}
4 >= 3
```

### Data types and basic statistics

There are many types of ways to store data in R, including:

-   vectors
-   lists
-   matrices
-   data frames

You can create a vector of any size using `c()`:

```{r, results='hide'}
y <- c(1, 2, 3, 4, 5)
```

You can perform basic mathematical operations of vectors using `sum()`, `mean()`, `sd()`, `var()`, `median()`, `IQR()`, `range()`, `quantile()`, `length()`:

```{r}
mean(y)
```

You can visualize simple plots of vectors using `hist()`, `barplot()`, `boxplot()`:

```{r}
boxplot(y)
```

A list is a set of vectors

```{r}
a <- list(c(1, 2, 3), c(3, 2, 1, 0, -1, -2, -3))
print(a)
```

Which is equivalent to making a dataframe of not necessarily the same lengths

```{r}
df <- data.frame(x = c(0, -1, 2), y = c(3, 2, 1))
df
```

### Accessing data

There are three main forms of accessing data:

-   `[ ]` to access a particular index of a vector
-   `[[ ]]` to access a particular index of a list
-   `$` to access an entire vector

```{r}
# get the third element of vector y
y <- c(1, 2, 3, 4, 5)
y[3]
```

```{r}
# get the third vector of list b
b <- list(c("Zebra"), c(1, 2, 3), c(-2, 0, 2, 4, 6))
b[[3]]
```

```{r}
# get the first column of dataframe df2
df2 <- data.frame(a = c("Apple", "Banana", "Chocolate"),
                  b = c(5, 10, 15))
df2$a
```

## Tidyverse code

While base R code is incredibly versatile, modern R users utilize the `tidyverse` package for data wrangling, visualization, and simple analysis, which comes with its own, slight variation on coding with R. This book will primarily use `tidyverse` for most code.

Let's install and load the `tidyverse` package.

```{r, eval = F}
install.packages("tidyverse")
```

```{r}
library("tidyverse")
```

The `tidyverse` is package that loads multiple packages that work in harmony with each other.

-   `dplyr` - for data wrangling and manipulation
-   `forcats` - for reordering factor variables
-   `ggplot2` - for data visualization
-   `lubridate` - used for handling date and time data
-   `purrr` - a functional programming toolkit
-   `readr` - for reading reading data files
-   `stringr` - for handling string variables
-   `tibble` - provides better formatting for data frames
-   `tidyr` - provides tools to handle tibble-formatted data frame

To demonstrate the basic features of these packages, we will use the `airquality` dataframe that is pre-loaded into the base R package `datasets`. We will store this dataframe under the name `aq_df`.

```{r}
aq_df <- datasets::airquality
```

### Creating tibbles with `tibble`

Let's view simply what the dataset looks like.

```{r}
head(aq_df)
```

We can see that that there are six columns related to ozone, solar radiation levels, wind spead, temperature, the month and day. We can view what the dataset type is using the `class()` function

```{r}
class(aq_df)
```

We can see that the dataset is stored in R as a "data.frame", not a "tibble". We can change this using the `as_tibble()` function. Reminder that in order to save this change in our local R environment, we have to reassign the command to `aq_df`.

```{r}
aq_df <- as_tibble(aq_df)
```

Now, we can check the class of the dataset to see if it is a tibble.

```{r}
class(aq_df)
```

We can indeed see that the class of "aq_df" is "tbl", analagous to "tibble".

### Data wrangling with `dplyr`

#### `arrange()`

We can sort the dataframe by a particular column(s) using the `arrange()` function. Let's do this by ascending temperature

```{r}
arrange(aq_df, Temp)
```

If we wanted to do this by descending temperature, we could insert a hyphen in front of the column name.

```{r}
arrange(aq_df, -Temp)
```

We can even sort by more than one column. Let's sort by day and then month.

```{r}
arrange(aq_df, Day, Month)
```

#### `select()`

We can keep or remove certain columns using the `select()` function. For example, let's say we know we will not use solar radiation for our particular analysis. We can remove it using a hyphen in front of the column name.

```{r}
select(aq_df, -Solar.R)
```

As you can see, the solar radiation column has been removed. Let's say we knew we only wanted to keep ozone levels, month, and day. We can specify that we only want to keep these columns as well.

```{r}
select(aq_df, Ozone, Month, Day)
```

#### `filter()`

Just as we can remove certain columns from our dataset, we can remove certain rows using the `filter()` function. Let's say we only cared about ozone levels in May. We can specify a boolean argument.

```{r}
filter(aq_df, Month == 5)
```

Notice that we are down to 31 rows from the original 153 rows now, since all the data outside of May has been removed.

#### `mutate()`

We can make new columns using the `mutate()` function. Let's say we are interested in creating an indicator variable for days when the ozone levels are above a certain threshold, say, 30. We can use the `ifelse()` function within the `mutate()` function to set the row to the value 1 if the ozone level is \>30, and 0 otherwise. We will name it "Ozone_above_30".

```{r}
mutate(aq_df, Ozone_above_30 = ifelse(Ozone > 30, 1, 0))
```

### The pipe operator: `%>%`

Let's say we wanted to arrange, select, filter, and mutate our dataset all in one line of code. In order to achieve this, we could do something like the following.

```{r}
mutate(filter(select(arrange(aq_df, -Temp), -Solar.R), Month == 5), Ozone_above_30 = ifelse(Ozone > 30, 1, 0))
```

As you can see from the code above, this operation is difficult to read and difficult to debug if there is an error in the code. For example, try to find what is wrong with this line of code.

```{r, eval = F}
mutate(filter(select(arrange(aq_df, -Temp), -Solar.R) Month == 5), Ozone_above_30 = ifelse(Ozone > 30, 1, 0))
```

You may have quickly noticed that the comma after the \`select(..., -Solar.R)" function is missing, but on another day when you're tired, you might miss this error and spend 30 minutes slamming your head onto your desk wondering why your code doesn't run even though it looks perfectly fine. Also, many R error messages are not very accurate. For example, this error says "unexpected symbol" when the problem is that it's missing a symbol (i.e., the comma).

Additionally, the nested functions mean that the first operation (i.e., `arrange(aq_df, -Temp)`) is in the middle of the line of code. Similarly, `mutate()` is at the front despite being the last operation. This makes it tricky to quickly understand the order of operations.

The pipe operator solves this problem. The pipe operator allows you to separate separate commands in a sequential manner while retaining the ability to run the entire operation in a single command. The above code is analagous to the following:

```{r}
arrange(aq_df, -Temp) %>%
  select(-Solar.R) %>%
  filter(Month == 5) %>%
  mutate(Ozone_above_30 = ifelse(Ozone > 30, 1, 0))
```

As you can see, using `%>%` allows us to first specify that we would like to arrange the data. Then, after calling the operator, we can state that we would like to remove solar radiation. Importantly, we don't have to re-specify the dataset argument since the pipe operator carries over the dataset from the previous operation. Then, the dataset is filtered, and finally mutated.

I hope you can see that the code with the pipe operator is more digestible and easier to debug if there was an error. We will code using the pipe operator for the rest of the book.

### Making graphs with `ggplot2`

The best package (in our opinion) for visualization is the `ggplot2` package.

Briefly, there are four main steps to a `ggplot2` graph:

-   Specify the dataset
-   Initialize the visualization using the `ggplot()` function
-   Specify the aesthetics using `aes()` (aka, your `x`, `y`, and other aesthetics like `size` and `shape`)
-   Specify the graph type using `geom_*()` (aka, the kind of graph you want to make, such as scatterplot)

For example, using our `aq_df` dataset, we can create a scatterplot to explore whether temperature is associated with ozone levels.

```{r, warning=FALSE}
aq_df %>%
  ggplot(aes(x = Temp, y = Ozone)) +
  geom_point()
```

It looks like it is! Let's break down the four steps used in the example code above:

-   Specify the dataset: `aq_df`
-   Initialize the visualization: `ggplot()`
-   Specify the aesthetics: `aes(x = Temp, y = Ozone)`
    -   In this case, we have two (x and y), which we define within the `ggplot()` function using the `aes()` function
-   Specify the graph type: `geom_point()`
    -   This is the syntax to create a scatterplot

::: {.callout-important icon="false"}
To bind together the `ggplot()` function with the `geom_point()` function, the `+` operator is used rather than the `%>%` operator.
:::

We won't go much much deeper into `ggplot2` here, as the ability to make graphs in R is wide and deep. It would not be possible to cover all of the possibilities just on this page. A great resource to learn more about `ggplot2` package can be found here: <https://rstudio.github.io/cheatsheets/html/data-visualization.html>

We will cover different visualizations relevant to exposure assessment throughout the semester that will build your `ggplot2` toolkit.
